Detecting Treatment-Subgroup Interactions in Clustered Data with Generalized Linear Mixed-Effects Model Trees

### 2015-09-20 pkg/glmertree/paper/treatment_interactions.Rnw
- Removed frequency tables for tree sizes - Shortened some of the subsections somewhat
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@793 255f4249-2263-4417-b1ea-8498779e27a7

\section{Abstract}
Identification of subgroups of patients for which treatment A is more effective than treatment B, and vice versa, is of key importance to the development of personalized medicine. Several tree-based algorithms have been developed for the detection of such treatment-subgroup interactions. In many instances, however, datasets may have a clustered structure, where observations are clustered within, for example, research centers, studies or persons. In the current paper we propose a new algorithm, generalized linear mixed-effects model (GLMM) trees, that allows for detection of treatment-subgroup interactions, as well as estimation of cluster-specific random effects. The algorithm uses model-based recursive partitioning (MOB) to detect treatment-subgroup interactions, and GLMMs for estimation of random-effects parameters. In a simulation study, we evaluate the performance of GLMM trees and compare it with that of MOB trees without random-effects estimation. In datasets without treatment-subgroup interactions, GLMM tree was found to have a much lower Type I error rate than MOB trees without random effects (4\% and 33\%, respectively). Furthermore, in datasets with treatment-subgroup interactions, GLMM trees recovered the true treatment subgroups much more often than MOB without random effects (90\% and 61\% of the datasets, respectively). Also, GLMM trees predicted treatment outcome differences more accurately than MOB trees without random effects (average accuracy of .94 and .88, respectively). We illustrate the application of GLMM tree on a patient-level dataset of a meta-analysis on the effects of psycho- and pharmacotherapy for depression. We conclude that GLMM trees are a promising algorithm for the detection of treatment-subgroup interactions in clustered datasets, and discuss some directions for future research.\\
\\
\textit{Keywords}: model-based recursive partitioning, treatment-subgroup interactions, random effects, generalized linear mixed-effects model, classification and regression trees

@@

In research assessing the efficacy of treatments for somatic and psychological disorders, the one-size-fits-all paradigm is slowly losing ground, and stratified medicine is becoming increasingly important. Stratified medicine presents the challenge of finding which patients respond best to which treatments. This can be referred to as the detection of treatment-subgroup interactions \cite<e.g., >{DoovyDuss14}. In most cases, treatment-subgroup interactions are studied using linear models, such as factorial analysis of variance techniques, in which potential moderators have to be specified a-priori, have to be checked one at a time, and continuous moderator variables have to be discretized a-priori. This may hamper identification of which treatments work best for whom, especially when there are no a-priori hypotheses about treatment-subgroup interactions. As noted by \citeA{KreayFran06}, there is a need for methods that generate, instead of test, hypotheses and that are specifically directed at the detection of treatment interactions.

Tree-based methods are such hypothesis-generating methods, as they can automatically detect subgroups which differ in the expected outcomes for one or more treatments. Due to their flexibility, tree-based methods are preeminently suited to the detection of treatment-subgroup interactions: they can handle many potential predictor variables at once and can automatically detect (higher order) interactions between predictor variables \cite{StroyMall09}. Several promising tree-based algorithms for the detection of treatment-subgroup interactions have been developed (e.g., \citeNP{DussyMech14, DussyMeul04, SuyTsai09, FostyTayl11, LipkyDmit11, ZeilyHoth08}; see \citeNP{DoovyDuss14} for an overview). Among these methods, model-based recursive partitioning (MOB; \citeNP{ZeilyHoth08}) seems to be the most flexible tool for detecting treatment-subgroup interactions as it offers a generic inferential framework that can be coupled with a broad range of parametric modeling strategies fitted by M-type estimators \cite{ZeilyHoth08}. Specifically, this model class encompasses the generalized linear model (GLM) and GLM-based MOB trees have been successfully applied by \citeA{DrieySmit14} in the detection of subgroups with differential treatment outcomes for two different psychotherapies.

However, none of the aforementioned tree-based algorithms allow for taking into account the clustered structure of many datasets. In such cases, researchers may want to detect treatment-subgroup interactions in datasets with a clustered structure \cite<e.g., >{KoopyHeij07}. For example, in individual-level patient data meta-analyses, in which datasets of multiple trials evaluating the effects of the same treatments are pooled. In such analyses, the clustered structure of the dataset should be taken into account by including study-specific effects in the model, prompting the need for modeling random effects \cite<e.g., >{CoopyPatt09,HiggyWhit01}. Likewise, longitudinal datasets, and datasets from multi-center trials typically also require modeling of random effects. Ignoring the clustered structure of datasets may lead to biased inference due to underestimated standard errors \cite<e.g., >{BrykyRaud92,NooryOpde05}. More specifically, when the interest is in subgroup detection, ignoring random effects on the outcome variable may result in the detection of spurious subgroups \cite<e.g., >{SelaySimo12}.

@@

In what follows, we will introduce the existing frameworks for estimating treatment effects: the generalized linear model (GLM), model-based recursive partitioning (MOB), and the generalized linear mixed-effects model (GLMM). Then, we introduce our new algorithm, which combines MOB and the GLMM: generalized linear mixed-effects model trees. Subsequently, we evaluate the performances of the GLMM tree algorithm in simulated datasets, and illustrate application of the algorithm with an exisiting dataset of a patient-level meta-analysis on the effects of treatments for depression by \citeA{CuijyWeit14}.
%
Before we discuss the methods for estimating treatment effects, we will introduce an artificial motivating data set, with which the methods will be illustrated. After we introduce the GLMM tree algorithm, we present a simulation study, in which we evaluate GLMM trees comparative accuracy. Finally, in the application, we use GLMM tree to detect treatment-subgroup interactions in an existing dataset on the effects of treatments for depression.

@@

To illustrate the application of the methods to be discussed, we will use a simulated dataset of 150 observations, which were randomly assigned to Treatment~1 or Treatment~2. Every observation has a value for the response variable, with which the effect of treatment is assessed: the posttreatment total score on a depression inventory. Further, all observations have values for three covariates: duration of depressive symptoms prior to treatment in months (range 0--15); age in years (range 18--75); anxiety inventory total score (range 3--18).

@@

The algorithm initializes by setting all $b$ values to $0$, since the random-effects (and also the fixed-effects) parts are initially unknown. In every iteration, the GLM tree and random-effects coefficients $b$ are re-estimated. The GLM tree is estimated, given the estimated $\hat{b}$ from the last iteration, and the $b$ values are estimated, given the estimated GLM tree from the current iteration. Iterations are continued until convergence, which is monitored by computing the log-likelihood criterion of the mixed-effects model in Equation~\ref{eq:mixedeffects}.

@@

\section{Emperical evaluation}

We will asses the performance of GLMM tree in recovering treatment-subgroup interactions, and predicting differences between the outcomes of two treatments, in simulated datasets with continuous outcomes. In addition, we will compare the performance of GLMM tree with that of GLM tree. In the simulation study, the main interest will be in the effects of sample size, and the presence and magnitude of treatment-subgroup interactions and random effects, but other parameters will be varied, as well.

For GLMM tree, we expect the accuracy of recovered trees and predictions to improve with increasing sample size, and magnitude of the differences in treatment outcomes. For GLM tree, we have the same expectation, when random effects are absent; that is, when the variance of the random coefficients is zero, we expect GLM tree and GLMM tree to perform equally well. When random effects are present, we expect GLMM tree to perform better than GLM tree, and more so when the variance of random-effects coefficients is larger.

@@

\subsubsection{Datasets with treatment-subgroup interactions}
For generating datasets with treatment-subgroup interactions, we used a treatment-subgroup interaction design from \citeA{DussyMech14}, which is also depicted in Figure~\ref{fig:modelC}. Figure~\ref{fig:modelC} shows two subgroups with mean differences in treatment outcomes, and two subgroups without mean differences in treatment outcomes. The four subgroups are characterized by their values on the partitioning variables $U_2$, and $U_1$ or $U_5$. In other words, $U_1$, $U_2$ and $U_5$ are true partitioning variables, whereas the other potential partitioning variables ($U_3$, $U_4$, $U_6$ through $U_{15}$) are noise variables.

@@

\begin{enumerate} 
\item Three levels for the total number of observations: $N=200$, $N=500$, $N=1000$.
\item Two levels for the number of potential partitioning covariates $U_1$ through $U_K$: $K=5$, $K=15$ (where only $U_1$, $U_2$ and $U_5$ are true partitioning variables).
\item Two levels of intercorrelations between the covariates $U_1$ through $U_K$: $\rho_{U_k,U_{k'}}=0.0$, $\rho_{U_k,U_{k'}}=0.3$.
\item Three levels for the number of clusters: $M=5$, $M=10$, $M=25$.
\item Three levels for the population standard deviation of the normal distribution from which the cluster specific intercepts are drawn: $\sigma_{b}=0$, $\sigma_{b}=5$, $\sigma_{b}=10$.
\item Three levels for the intercorrelations between $b$ and one of the $U_k$ variables: $b$ and $U_k$ uncorrelated, $b$ correlated with a true partitioning covariate (i.e., $U_2$, $U_1$, or $U_5$, introducing a correlation of about 0.42), $b$ correlated with a non-partitioning covariate (i.e., $U_3$ or $U_4$, introducing a correlation of about 0.42). 
\item Two different levels for $\beta_{j1}$, the unstandardized mean difference in treatment outcomes, in subgroups with differential treatment effects. The levels for mean differences in subgroups with differential treatment effect were $|\beta_{j1}| = 2.5$ (corresponding to a medium effect size, Cohen's $d = 0.5$; \citeNP{Cohe92}) and $|\beta_{j1}| = 5.0$ (corresponding to a large effect size; Cohen's $d = 1.0$).
\end{enumerate}

@@

It should be noted that when $\sigma_b = 0$, the correlation between $b$ and one of the $U_k$ variables is 0, by definition. However, datasets were created for this condition, to allow for a full factorial design of the simulation study; in reality, $b$ and $U$ are uncorrelated in these instances.

@@

To generate the cluster-specific intercepts $b_m$, we partitioned the sample into equally-sized clusters, conditional on one of the variables $U_1$ through $U_5$, producing the correlations in the sixth facet of the simulation design. For each cluster we drew a single $b_m$ from a normal distribution with mean 0 and the value of $\sigma_{b}$ given by the fifth facet of the simulation design. When $b$ was correlated with one of the potential partitioning variables, the correlated potential partitioning variable was randomly selected.

@@

For every dataset, accuracy and size of the GLM and GLMM tree was evaluated. We calculated the total number of nodes in every tree, and compared it with the true tree size. For datasets without treatment-subgroup interactions, this allowed us to assess tree accuracy in terms of Type I error: the probability that the dataset is erroneously partitioned. For datasets with treatment-subgroup interactions, this allowed us to assess the probability that the dataset is erroneously not partitioned, and the extent to which the algorithms may detect spurious subgroups.

For datasets with treatment-subgroup interactions, we assessed the accuracy of the trees created by GLM and GLMM tree. An accurately recovered tree was defined as a tree with (1) the true tree size (i.e., total number of nodes equals 7), (2) the first split in the tree involving variable $U_2$ and a value of $30 \pm 5$, (3) the next split on the left involving variable $U_1$ and a value of $17 \pm 5$, and (4) the next split on the right involving variable $U_5$ and a value of $63 \pm 5$. Note that the allowance of $\pm 5$ equals an allowance of plus or minus half the population standard deviation of the partitioning variable ($\sigma_{U_k}$).

To assess the impact of data-generating parameters on tree size for both algorithms, we performed ANOVAs with algorithm type and the parameters of the data-generating process as independent variables. In addition, interactions between algorithm type and each of the data-generating parameters were also entered as independent variables. The impact of predictors with main and/or interaction effects which explained a proportion of $> .01$ of variance were further investigated using graphical displays.

To assess the impact of data-generating parameters on tree accuracy in datasets with treatment-subgroup interactions, we used a GLM with algorithm type and the parameters of the data-generating process as independent variables. In addition, interactions between algorithm type and each of the data-generating parameters were also entered as independent variables. The impact of predictors with main and/or interaction effects with unstandardized regression coefficients $> .5$ (i.e., an in- or decrease in the log-odds of .5) were further investigated using graphical displays.

@@

We evaluated predictive accuracy of GLM and GLMM trees by calculating correlations between true and predicted treatment-effect differences ($\beta_{j1}$ in Figure~\ref{fig:modelC}) for test observations. Note that this correlation was only assessed for datasets with treatment-subgroup interactions, as the true treatment differences have a constant value in datasets without treatment-subgroup interactions.

@@

For every dataset, correlation coefficients for each algorithm were calculated, representing the linear association between the true and predicted treatment-effect differences. To assess the impact of data-generating parameters on predictive accuracy, we performed ANOVAs with algorithm type and the parameters of the data-generating process as independent variables. In addition, interactions between algorithm type and each of the data-generating parameters were also entered as independent variables. The impact of predictors with main and/or interaction effects which explained a proportion of $> .01$ of variance were further investigated using graphical displays.

@@

R \cite{R14} was used for generation and analysis of all datasets. The \verb|partykit| package (version 1.0-2; \citeNP{HothyZeil15}) was employed for estimating GLM trees using the \verb|lmtree| function for normal linear regressions and \verb|glmtree| would be available for GLM trees with other response distributions. For estimation of GLMMs the \verb|lmer| (or \verb|glmer|, respectively) from the \verb|lme4| package (version 1.1-7; \citeNP{BateyMeac12}) was employed using restricted maximum likelihood (REML) estimation.

@@

In all applications, the significance level $\alpha$ for the parameter instability tests in the trees was set to .05 with a Bonferroni correction applied for multiple testing. The minimum number of observations per node in the tree was set to 20 and the maximum tree depth was set to four, thus limiting the number of potential subgroups to eight. The iteration in the GLMM tree algorithm was repeated until the log-likelihood changes between two subsequent iterations fell below .001.

@@

In Table~\ref{tab:treesize_main}, tree sizes for GLM and GLMM trees for datasets without treatment-subgroup interactions are presented. Overall, smaller trees were created by GLMM tree: the average tree size was 1.09 ($\mathrm{SD}=0.44$) for GLMM tree, and 2.02 ($\mathrm{SD}=1.68$) for GLM tree. The estimated probability that a dataset was erroneously partitioned was very small for GLMM tree (.04; Table~\ref{tab:treesize_main}), and much larger for GLM tree (.33; Table~\ref{tab:treesize_main}).

\begin{table}[t!]
	\caption{Tree size distributions for GLM and GLMM tree for datasets without treatment-subgroup interactions.}
	\small
	\begin{tabular}{lrrrrrrr}
		\thickline
		&\multicolumn{7}{c}{tree size}\\
		\cline{2-7}
		&   1    &   3   &   5   &   7   &   9   &  11  &   total \\
		\hline
		GLMM tree	& 20625 & 932 & 43 &0&0&0& 21,600\\
			  		& (.96) & (.04) & ($<.01$) &(.00)&(.00)&(.00)&(1.00) \\
   		GLM tree		& 14501  & 4202 & 2013 & 802 & 79 & 3 & 21,600 \\ 
    			& (.67)  & (.19) & (.09) & (.04) & ($<.01$) & ($<.01$) & (1.00)\\ 
		\hline
  		\multicolumn{8}{l}{\textit{Note. }Bracketed values are proportions. Tree sizes are expressed as the total}\\
  		\multicolumn{8}{l}{number of nodes in a tree. A tree with a total of $J$ nodes has $(J+1)/2$ }\\
  		\multicolumn{8}{l}{terminal nodes; the true tree size in datasets without treatment-subgroup}\\	  	\multicolumn{8}{l}{interactions was 1.}\\
	\end{tabular}
	\label{tab:treesize_main}
\end{table}

@@

A graphical display was used to asses the effects of sample size, $\sigma_b$ and the correlation between $b$ and one of the $U_k$ variables, on tree size (Figure~\ref{fig:xyplot_treesize_nointeract}). When random effects were absent (i.e., $\sigma_{b}=0$), both GLM and GLMM tree tend to create trees of size 1. In the presence of random effects, GLMM tree also tends to create trees of size 1, but GLM tree created much larger trees, when $b$ was correlated to one of the $U_k$ variables. This effect was stronger when sample size was larger.

@@

In datasets with treatment-subgroup interactions, GLMM trees were also smaller than GLM trees. For these datasets, the true tree size was 7 (4 terminal nodes and 3 inner nodes; Figure~\ref{fig:modelC}). Th distribution of tree sizes for GLM and GLMM tree in datasets with treatment-subgroup interactions are presented in Table~\ref{tab:treesize_interactions}. The average size of GLMM trees was 7.16 ($\mathrm{SD}=0.62$), and the average size of GLM trees was 8.12 ($\mathrm{SD}=2.05$). The estimated probability that a dataset was erroneously not partitioned was 0, for both GLM and GLMM tree. However, Table~\ref{tab:treesize_interactions} shows that a proportion of .91 of GLMM trees matched the true tree size, whereas a proportion of only .63 of GLM trees matched the true tree size (Table~\ref{tab:treesize_interactions}).

\begin{table}[t!]
\caption{Tree size distributions for GLM and GLMM tree for datasets with treatment-subgroup interactions.}
\small
	\begin{tabular}{lrrrrrrrr}
		\thickline
		&&\multicolumn{7}{c}{tree size}\\
		\cline{2-8}
		&     3   &   5   &   7   &   9   &  11  &   13   &   15  & total \\
		\hline
	  	GLMM tree	& 2 & 281& 29358 & 2675 & 83 & 1 &0& 32,400 \\
					&  ($<.01$)  & ($<.01$)& (.91) & ($.08$) & ($<.01$) & ($<.01$) &(.00)& (1.00) \\

	  	GLM tree	& 139 & 955 & 20530 & 4565 & 4164 & 1619 & 428 & 32,400\\ 
	 			 	& ($<.01$) & (.03) & (.63) & (.14) & (.13) & (.05) & (.01) & (1.00) \\ 
		\hline
  		\multicolumn{9}{l}{\textit{Note. } Bracketed values are proportions. Tree sizes are expressed as the total number of}\\
  		\multicolumn{9}{l}{nodes in the tree. A tree with a total of $J$ nodes has $(J+1)/2$ terminal nodes; the true}\\	
  		\multicolumn{8}{l}{tree size in datasets with treatment-subgroup interactions was 7.}\\
	\end{tabular}
	\label{tab:treesize_interactions}
\end{table}

A graphical display was used to assess the effects of sample size, $\sigma_b$ and the correlation between $b$ and one of the $U_k$ variables, on tree size (Figure~\ref{fig:xyplot_treesize_interact}). When random effects were absent (i.e., $\sigma_{b}=0$), both GLM and GLMM tree created trees with a size of about 7, on average.

@@

Clear differences in performance between GLM and GLMMtree were observed when $\sigma_b > 0$. When $b$ is not correlated with one of the $U_k$ variables, when sample size is small (i.e., 200) and when $\sigma_b$ is large (i.e., 10), GLM tree has difficulty detecting splits and grows trees that are too small, on average. When $b$ is not correlated with one of the $U_k$ variables and when sample size is larger (i.e., 500 or 1000), GLM and GLMM trees are about the same size (i.e., $\approx 7$). When $b$ is correlated with one of the $U_k$ variables, GLM creates spurious splits, especially when sample size is larger (i.e., 500 or 1000) and when $\sigma_b$ is large (i.e., 10). This effect was stronger when $b$ was correlated to a non-splitting variable.

@@

To assess the accuracy of the trees created by GLM and GLMM tree, we inspected the variables and values that were selected for partitioning in every dataset. For the first split, GLMM tree always selected the true partitioning variable ($U_2$). GLM tree selected a wrong partitioning variable (I.e., $U_1$ in only one dataset. The true splitting value for $U_2$ was 30 (Figure~\ref{fig:modelC}), and the mean splitting value selected for the first split (involving $U_2$) was 29.94, for both GLM and GLMM tree. However, GLM tree showed somewhat higher variability in recovering the splitting value for the first split (involving $U_2$), than did GLMM tree ($\mathrm{SD}=0.154$ and $\mathrm{SD}=0.127$, respectively).

>>>>

\section{Abstract}
Identification of subgroups of patients for which treatment A is more effective than treatment B, and vice versa, is of key importance to the development of personalized medicine. Several tree-based algorithms have been developed for the detection of such treatment-subgroup interactions. In many instances, however, datasets may have a clustered structure, where observations are clustered within, for example, research centers, studies or persons. In the current paper we propose a new algorithm, generalized linear mixed-effects model (GLMM) trees, that allows for detection of treatment-subgroup interactions, as well as estimation of cluster-specific random effects. The algorithm uses model-based recursive partitioning (MOB) to detect treatment-subgroup interactions, and a GLMM for the estimation of random-effects parameters. In a simulation study, we evaluate the performance of GLMM tree and compare it with that of MOB without random-effects estimation. GLMM tree was found to have a much lower Type I error rate than MOB trees without random effects (4\% and 33\%, respectively). Furthermore, in datasets with treatment-subgroup interactions, GLMM tree recovered the true treatment subgroups much more often than MOB without random effects (in 90\% and 61\% of the datasets, respectively). Also, GLMM tree predicted treatment outcome differences more accurately than MOB without random effects (average predictive accuracy of .94 and .88, respectively). We illustrate the application of GLMM tree on a patient-level dataset of a meta-analysis on the effects of psycho- and pharmacotherapy for depression. We conclude that GLMM tree is a promising algorithm for the detection of treatment-subgroup interactions in clustered datasets.\\
\\
\textit{Keywords}: model-based recursive partitioning, treatment-subgroup interactions, random effects, generalized linear mixed-effects model, classification and regression trees

@@

In research assessing the efficacy of treatments for somatic and psychological disorders, the one-size-fits-all paradigm is slowly losing ground, and personalized medicine is becoming increasingly important. Stratified medicine presents the challenge of finding which patients respond best to which treatments. This can be referred to as the detection of treatment-subgroup interactions \cite<e.g., >{DoovyDuss14}. In most cases, treatment-subgroup interactions are studied using linear models, such as factorial analysis of variance techniques, in which potential moderators have to be specified a-priori, have to be checked one at a time, and continuous moderator variables have to be discretized a-priori. This may hamper identification of which treatments work best for whom, especially when there are no a-priori hypotheses about treatment-subgroup interactions. As noted by \citeA{KreayFran06}, there is a need for methods that generate, instead of test, hypotheses and that are specifically directed at the detection of treatment interactions.

Tree-based methods are such hypothesis-generating methods, as they can automatically detect subgroups which differ in the expected outcomes for one or more treatments. Due to their flexibility, tree-based methods are preeminently suited to the detection of treatment-subgroup interactions: they can handle many potential predictor variables at once and can automatically detect (higher order) interactions between predictor variables \cite{StroyMall09}. Several promising tree-based algorithms for the detection of treatment-subgroup interactions have been developed (e.g., \citeNP{DussyMech14, DussyMeul04, SuyTsai09, FostyTayl11, LipkyDmit11, ZeilyHoth08}; see \citeNP{DoovyDuss14} for an overview). Among these methods, model-based recursive partitioning (MOB; \citeNP{ZeilyHoth08}) seems to be the most flexible tool for detecting treatment-subgroup interactions, as it offers a generic inferential framework that can be coupled with a broad range of parametric modeling strategies fitted by M-type estimators \cite{ZeilyHoth08}. Specifically, this model class encompasses the generalized linear model (GLM). GLM-based MOB has been successfully applied by \citeA{DrieySmit14} in the detection of subgroups of patients with depression, which differ in the effect of two different psychotherapies. Also, \citeA{SeibyZeil15} applied MOB for in the detection of subgroups of patients with amyotrophic lateral sclerosis (ALS), which differ in the effect of treatment with Riluzole.

In many cases, researchers may want to detect treatment-subgroup interactions in datasets with a clustered structure. However, none of the aforementioned tree-based algorithms allow for taking into account the clustered structure of a dataset. For example, in individual-level patient data meta-analyses, in which datasets of multiple trials evaluating the effects of the same treatments are pooled \cite<e.g., >{KoopyHeij07}. In such analyses, the clustered structure of the dataset should be taken into account by including study-specific effects in the model, prompting the need for modeling random effects \cite<e.g., >{CoopyPatt09,HiggyWhit01}. Likewise, longitudinal datasets, and datasets from multi-center trials typically also require modeling of random effects. Ignoring the clustered structure of datasets may lead to biased inference due to underestimated standard errors \cite<e.g., >{BrykyRaud92,NooryOpde05}. More specifically, when the interest is in subgroup detection, ignoring random effects on the outcome variable may result in the detection of spurious subgroups \cite<e.g., >{SelaySimo12}.

@@

In what follows, we will introduce the existing frameworks for estimating treatment effects: the generalized linear model (GLM), model-based recursive partitioning (MOB), and the generalized linear mixed-effects model (GLMM). Then, we introduce our new algorithm, which combines MOB and the GLMM: generalized linear mixed-effects model trees. Subsequently, in the Empirical Evaluation section, we evaluate the performance of the GLMM tree algorithm in simulated datasets. In the Application section, we illustrate the application of the algorithm, using an existing dataset of a patient-level meta-analysis on the effects of psycho- and pharmacotherapeutic treatments for depression of \citeA{CuijyWeit14}. But before we discuss the frameworks for estimating treatment effects, we will introduce an artificial motivating data set with which the frameworks and methods to be discussed, will be illustrated.

@@

We created a simulated dataset of 150 observations, which were randomly assigned to Treatment~1 or Treatment~2. Every observation has a value for the response variable, with which the effect of treatment is assessed: the posttreatment total score on a depression inventory. Further, all observations have values for three covariates: duration of depressive symptoms prior to treatment in months (range 0--15); age in years (range 18--75); anxiety inventory total score (range 3--18).

@@

The algorithm initializes by setting all $b$ values to $0$, since the random-effects (and also the fixed-effects) parts are initially unknown. In every iteration, the GLM tree and random-effects coefficients $b$ are re-estimated. The GLM tree is estimated, given the estimated $\hat{b}$ from the last iteration, and the $b$ values are estimated, given the estimated GLM tree from the current iteration. Iterations are continued until convergence, which is monitored by computing the log-likelihood criterion of the mixed-effects model in Equation~\ref{eq:glimmertree}.

@@

\section{Empirical evaluation}

We will asses the performance of GLMM tree in recovering treatment-subgroup interactions, and predicting differences between the outcomes of two treatments, in simulated datasets with continuous outcomes. In addition, we will compare the performance of GLMM tree with that of GLM tree. In the simulation study, our main interest will be in the effects of sample size, the presence and magnitude of treatment-subgroup interactions, and the presence and magnitude of the random effects, but additonal parameters will be varied as well.

When random effects are absent from the datasets, we expect GLM and GLMM tree to perform equally well. In the presence of random effects, we expect GLMM tree to outperform GLM tree. We expect the difference in performance between both algorithms to increase, with increasing differences in treatment outcomes, increasing variance of random-effects coefficients, and/or increasing sample sizes.

@@

\subsubsection{Datasets with treatment-subgroup interactions}
For generating datasets with treatment-subgroup interactions, we used a treatment-subgroup interaction design from \citeA{DussyMech14}, which is also depicted in Figure~\ref{fig:modelC}. Figure~\ref{fig:modelC} shows two subgroups with mean differences in treatment outcomes, and two subgroups without mean differences in treatment outcomes. The four subgroups are characterized by their values on the partitioning variables $U_2$, and $U_1$ or $U_5$. That is, $U_1$, $U_2$ and $U_5$ are true partitioning variables, whereas other potential partitioning variables ($U_3$, $U_4$, $U_6$ through $U_{15}$) are noise variables.

@@

\begin{enumerate} 
\item Three levels for the total number of observations: $N=200$, $N=500$, $N=1000$.
\item Two levels for the number of potential partitioning covariates $U_1$ through $U_K$: $K=5$, $K=15$ (where only $U_1$, $U_2$ and $U_5$ are true partitioning variables).
\item Two levels of intercorrelations between the covariates $U_1$ through $U_K$: $\rho_{U_k,U_{k'}}=0.0$, $\rho_{U_k,U_{k'}}=0.3$.
\item Three levels for the number of clusters: $M=5$, $M=10$, $M=25$.
\item Three levels for the population standard deviation of the normal distribution from which the cluster specific intercepts are drawn: $\sigma_{b}=0$, $\sigma_{b}=5$, $\sigma_{b}=10$.
\item Three levels for the intercorrelations between $b$ and one of the $U_k$ variables: $b$ and $U_k$ uncorrelated, $b$ correlated with a true partitioning variable (i.e., $U_2$, $U_1$, or $U_5$, introducing a correlation of $\approx 0.42$), $b$ correlated with a non-partitioning covariate (i.e., $U_3$ or $U_4$, introducing a correlation of $\approx 0.42$)\footnote{Note that, when $\sigma_b = 0$, the correlation between $b$ and one of the $U_k$ variables is 0, by definition. However, datasets were created for this condition, to allow for a full factorial design of the simulation study; in reality, $b$ and $U$ are uncorrelated in these instances.}. 
\item Two different levels for $\beta_{j1}$, the unstandardized mean difference in treatment outcomes, in subgroups with differential treatment effects. The levels for mean differences in subgroups with differential treatment effect were $|\beta_{j1}| = 2.5$ (corresponding to a medium effect size, Cohen's $d = 0.5$; \citeNP{Cohe92}) and $|\beta_{j1}| = 5.0$ (corresponding to a large effect size; Cohen's $d = 1.0$).
\end{enumerate}

@@

@@

To generate the cluster-specific intercepts $b_m$, we partitioned the sample into equally-sized clusters, conditional on one of the variables $U_1$ through $U_5$, producing the correlations in the sixth facet of the simulation design. For each cluster, we drew a single value $b_m$ from a normal distribution with mean 0 and the value of $\sigma_{b}$ given by the fifth facet of the simulation design. When $b$ was correlated with one of the potential partitioning variables, the correlated potential partitioning variable was randomly selected.

@@

For every dataset, the total number of nodes in the resulting GLM and GLMM tree were calculated. For datasets without treatment-subgroup interactions, this allowed us to assess tree accuracy in terms of Type I error: the probability that the dataset is erroneously partitioned (i.e., a tree of size $>1$ is created). For datasets with treatment-subgroup interactions, this allowed us to assess the probability that the dataset is erroneously not partitioned, and the extent to which the algorithms may detect spurious subgroups (i.e., a tree of size $> 7$ is created).

For datasets with treatment-subgroup interactions, we assessed the accuracy of the GLM and GLMM trees. An accurately recovered tree was defined as a tree with (1) the true tree size (i.e., tree size $= 7$), (2) the first split in the tree involving variable $U_2$ and a value of $30 \pm 5$, (3) the next split on the left involving variable $U_1$ and a value of $17 \pm 5$, and (4) the next split on the right involving variable $U_5$ and a value of $63 \pm 5$. Note that the allowance of $\pm 5$ equals an allowance of plus or minus half the population standard deviation of the partitioning variable ($\sigma_{U_k}$).

To assess the effects of the data-generating parameters on tree size for both algorithms, we performed ANOVAs with algorithm type and the parameters of the data-generating process as independent variables. In addition, interactions between algorithm type and each of the data-generating parameters were also entered as independent variables. The impact of predictors with main and/or interaction effects which explained a proportion of $> .01$ of variance was further assessed using graphical displays.

To assess the effects of the data-generating parameters on tree accuracy in datasets with treatment-subgroup interactions, we used a GLM with algorithm type and the parameters of the data-generating process as independent variables. In addition, interactions between algorithm type and each of the data-generating parameters were also entered as independent variables. The effects of predictors with main and/or interaction effects with unstandardized regression coefficients $> .5$ (i.e., an in- or decrease in the log-odds of .5) were further assessed using graphical displays.

@@

We evaluated predictive accuracy of GLM and GLMM trees by calculating correlations between the true and predicted treatment-effect differences ($\beta_{j1}$ in Figure~\ref{fig:modelC}). Note that this correlation was only assessed for datasets with treatment-subgroup interactions, as the true treatment differences have a constant value in datasets without treatment-subgroup interactions.

@@

For every dataset, correlation coefficients for each algorithm were calculated, representing the linear association between the true and predicted treatment-effect differences. To assess the effects of the data-generating parameters on predictive accuracy, we performed ANOVAs with algorithm type and the parameters of the data-generating process as independent variables. In addition, interactions between algorithm type and each of the data-generating parameters were also entered as independent variables. The effects of predictors with main and/or interaction effects which explained a proportion of $> .01$ of variance were further investigated using graphical displays.

@@

R \cite{R14} was used for generation and analysis of all datasets. The \verb|partykit| package (version 1.0-2; \citeNP{HothyZeil15}) was employed for estimating GLM trees using the \verb|lmtree| function for normal linear regressions. For other response distributions, the \verb|glmtree| function would be available. For estimation of GLMMs the \verb|lmer| (or \verb|glmer|, respectively) from the \verb|lme4| package (version 1.1-7; \citeNP{BateyMeac12}) was employed, using restricted maximum likelihood (REML) estimation.

@@

In all applications, the significance level $\alpha$ for the parameter instability tests in the trees was set to .05, with a Bonferroni correction applied for multiple testing. The minimum number of observations per node in the tree was set to 20 and the maximum tree depth was set to four, thus limiting the number of potential subgroups to eight. Iterations of the GLMM tree algorithm were repeated until the log-likelihood changes between two subsequent iterations fell below .001.

@@

Overall, smaller trees were created by GLMM tree: the average tree size was 1.09 ($\mathrm{SD}=0.44$) for GLMM tree, and 2.02 ($\mathrm{SD}=1.68$) for GLM tree. The estimated probability that a dataset was erroneously partitioned was .04 for GLMM tree, and .33 for GLM tree.

The effects of sample size, $\sigma_b$ and the correlation between $b$ and one of the $U_k$ variables on tree size were assessed with a graphical display (Figure~\ref{fig:xyplot_treesize_nointeract}). When random effects were absent (i.e., $\sigma_{b}=0$), both GLM and GLMM tree tended to create trees of size 1. In the presence of random effects, GLMM tree also tended to create trees of size 1, but GLM tree created much larger trees, when $b$ was correlated with one of the $U_k$ variables. This effect was stronger when sample size was larger.

@@

@@

In datasets with treatment-subgroup interactions, GLMM trees were also smaller than GLM trees. For these datasets, the true tree size was 7 (4 terminal nodes and 3 inner nodes; Figure~\ref{fig:modelC}). The average size of GLMM trees was 7.16 ($\mathrm{SD}=0.62$), and the average size of GLM trees was 8.12 ($\mathrm{SD}=2.05$). The estimated probability that a dataset was erroneously not partitioned was 0, for both GLM and GLMM tree. However, a proportion of .91 of GLMM trees matched the true tree size, whereas a proportion of only .63 of GLM trees matched the true tree size.

The effects of sample size, $\sigma_b$ and the correlation between $b$ and one of the $U_k$ variables on tree size were assessed with a graphical display (Figure~\ref{fig:xyplot_treesize_interact}). When random effects were absent (i.e., $\sigma_{b}=0$), both GLM and GLMM tree created trees with a size of about 7, on average. However, clear differences in performance between GLM and GLMMtree were observed when $\sigma_b > 0$. When $b$ is not correlated with one of the $U_k$ variables, when sample size is small (i.e., 200) and when $\sigma_b$ is large (i.e., 10), GLM tree has difficulty detecting splits and grows trees that are too small, on average. When $b$ is not correlated with one of the $U_k$ variables and when sample size is larger (i.e., 500 or 1000), GLM and GLMM trees are about the same size (i.e., $\approx 7$). When $b$ is correlated with one of the $U_k$ variables, GLM creates spurious splits, especially when sample size is larger (i.e., 500 or 1000) and when $\sigma_b$ is large (i.e., 10). This effect was stronger when $b$ was correlated to a non-splitting variable.

@@

@@

To assess the accuracy of the trees created by GLM and GLMM tree, we inspected the variables and values that were selected for partitioning in every dataset. For the first split, GLMM tree always selected the true partitioning variable ($U_2$). GLM tree selected a wrong partitioning variable (I.e., $U_1$ in only one dataset. The splitting value for $U_2$ selected for the first split was 29.94 for both GLM and GLMM tree, which is very close to the true splitting value of 30 (Figure~\ref{fig:modelC}). However, GLM tree showed somewhat higher variability in recovering the splitting value for the first split (involving $U_2$), than did GLMM tree ($\mathrm{SD}=0.154$ and $\mathrm{SD}=0.127$, respectively).
